### 2015-02-09 src/3_TimeSeries.tex
Cointegration example

In finance, a shock represents an unexpected change in a variable or a in its
error term in a particular time period. For stationary time series, shocks to
the system will gradually die away. That is, a shock during time $t$ will have a
smaller effect in time $t+1$, a smaller effect on $t+2$ and so on. For
non-stationary data, the persistence of shocks will always be infinite. The use
of non-stationary data can lead to spurious regressions. Spurious regression
dates back to Yule in 1926 \cite{yule1926} If two stationary
variables are generated as independent random series and are trending over time,
when one of those variables is regressed on the other, they could have a high
$R^2$ even if the two are totally unrelated. So, if standard regression
techniques are applied to non-stationary data, it could look good under standard
measures but valueless \cite{brooks2002}.

@@

\subsection{Integration}

@@

>>>>

In finance, a shock represents an unexpected change in a variable or a in its
error term in a particular time period. For stationary time series, shocks to
the system will gradually die away. That is, a shock during time $t$ will have a
smaller effect in time $t+1$, a smaller effect on $t+2$ and so on. For
non-stationary data, the persistence of shocks will always be infinite.

\subsection{Non-stationary processes}

There are different types of non-stationary time series models often found in economics:

\subsubsection{Deterministic trend}

Deterministic trend or trend stationary processes have the following form:

\[
X_t = f(t) + \epsilon{t} \quad ,
\]

\noindent where $t$ is the time trend and $\epsilon{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time:
\begin{itemize}
\item If $f(t) = \alpha + \beta t$ we have a linear trend model which is widely used. 
\item If $f(t)= \alpha \exp^rt$ we have an exponential growth curve.
\item If $f(t) = c_1 + c_2t + c_3 t^2$ we have a quadratic trend model
\item If $f(t) = \frac{1}{k+\alpha \beta ^t }$ we have a logistic curve
\end{itemize}

\subsubsection{Stochastic trend}
Stochastic trend processes are also called unit root or difference stationarity processes and have the following form:

\[
X_t = \mu + X_{t-1} + \epsilon_t
\]
\noindent where $\epsilon_t$ is a stationary process. When $\mu = 0 $ the process is called pure random walk and when $\mu \neq 0$  the process is called random walk with drift.

Alternatively this process can be expressed using the lag operator $L$ such as:

\[
(1-L) X_t = \mu  + \epsilon_t
\]

This process is also called unit root because the root of the characteristic equation ($1-z = 0$) is the unity.

A random walk with or without drift can be transformed to a stationary process by differencing the time serie once. The disadvantage of differencing is that the process loses one observation each time the time series is differentiated.

Apart from a stochastic trend, many economic financial time series seem to involve an exponential trend, this is the reason why researchers often take the logarithmic transformation before doing analysis.

Other less common forms of non-stationarity are structural break in mean and structural break in variance.

\newpage
\subsection{Spurious regression}
The use
of non-stationary data can lead to spurious regressions. Spurious regression
dates back to Yule in 1926 \cite{yule1926} If two stationary
variables are generated as independent random series and are trending over time,
when one of those variables is regressed on the other, they could have a high
$R^2$ even if the two are totally unrelated. So, if standard regression
techniques are applied to non-stationary data, it could look good under standard
measures but valueless \cite{brooks2002}.

@@

\subsection{Unit root tests}

\newpage
\subsection{Integration}

@@

\subsection{Johansen method}
Johansen in 1988 \cite{johansen1988} suggests a method for determining cointegration vectors.

There are two statistics for cointegration under Johansen approach: the trace (shown in equation \ref{eq:johansentrace}) and the maximum-eigenvalue statistic (shown in equation \ref{eq:johansenmax}).

\begin{equation}
\label{eq:johansentrace}
\lambda_{\text{trace}} (r) = -T \sum_{i=r+1}^g \ln(1-\hat{\lambda_i})
\end{equation}

\begin{equation}
\label{eq:johansenmax}
\lambda_{\text{max}} (r,r+1) = -T \ln(1-\hat{\lambda_{r+1}})
\end{equation}

Where $r$ is the number of cointegration vectors and $\hat]{\lambda}_i$ is the estimated value for the ith ordered eigenvalue.  $\lambda_{\text{trace}}$ null hypothesis is that the number of cointegration vectors is less than or equal to $r$ against that there are more than $r$. $\lambda_{\text{max}}$ has the null hypothesis that the number of cointegration vectors is $r$ against an alternative of $r
+1$.

\newpage
\texttt{Cointegration example}

If we have two-dimensional process $\mathbf{X}_t$, $t=1,\dots,T$ by:

\begin{eqnarray*}
\mathbf{X}_{1t} &=& \sum_{i=1}^t \epsilon_{1i} + \epsilon_{2t} \\
\mathbf{X}_{2t} &=& a \sum_{i=1}^t \epsilon_{1i} + \epsilon_{3t} 
\end{eqnarray*}

Since $\mathbf{X}_{1t}$ and $\mathbf{X}_{2t}$ are I(1) processes and there
exist a vector $\beta = [a -1]$ such that:

\[
\beta^\top \mathbf{X}_t = a \mathbf{X}_{1t} -\mathbf{X}_{2t} = 
a\epsilon_{2t} - \epsilon_{3t} \sim \text{I(0)}
\]

then, both processes are said to be cointegrated.

If we add a I(0) process
$\mathbf{X}_{3t} = \epsilon_{4t}$  we find that there exists two cointegration
vectors now: $\begin{bmatrix}a &-1& 0\end{bmatrix}$ and $\begin{bmatrix}0
&0&1\end{bmatrix}$ since:

\[
\beta^\top \mathbf{X}_t = 
\begin{bmatrix}
a & -1 & 0 \\
0 & 0 & 1
\end{bmatrix} 
\begin{bmatrix} 
\mathbf{X}_{1t} \\
\mathbf{X}_{2t} \\
\mathbf{X}_{3t}
\end{bmatrix} = 
\begin{bmatrix}
a\epsilon_{2t} - \epsilon_{3t} \\
\epsilon_{4t}
\end{bmatrix}
\]

\begin{figure}[!h]
  \centering
  \includegraphics[width=\textwidth]{img/coint-example}
  \caption{Cointegration example: two processes I(1)(blue and green) and one process I(0) (red)}
  \label{fig:cointex}
\end{figure}

This example shows how cointegration vectors describes the stable relations
between the processes by linear relations that are more stationary than the
original process. Cointegration vectors can be obtained using the Johansen method which gives with a certain probability the number of significant vectors. Figure \cite{fig:cointvectors1} shows the linear combinations obtained using cointegration vectors given by Johansen method. The method confirm that only two vectors are found and the third one given doesn't result in a I(0) process.

\begin{figure}[!h]
  \centering
 \begin{minipage}{\textwidth} %[b][\textheight][s]
  \centering
  \includegraphics[width=0.8\textwidth]{img/coint-example2} 
   \caption{Time series linear combination using the first cointegration vector }
   \includegraphics[width=0.8\textwidth]{img/coint-example3} 
    \caption{Time series linear combination using the second cointegration vector}
    \includegraphics[width=0.8\textwidth]{img/coint-example4}
 \caption{Time series linear combination using the third cointegration vector}
  \label{fig:cointvectors1}
\end{minipage}
\end{figure}

\newpage
